Group Theory in Physics (Course Note)

### 2016-11-13 Lecture note - Group Theory in Physics/sections/point-group.tex
Almost finised the Irreps if D_n and C_n. (representation, group theory)

\paragraph{Irreducible representation} Since it has $n$ elements and
$n$ conjugacy classes, with $\sum_{j=1}^n d_j^2 = n$, each irreducible
representation is $1$-dimensional. Let $R$ denotes the generator of
rotation by $2\pi/n$, then by calculating $D^j(R)^n=D^j(1)$, one can
easily find that for each irreducible representation $j$:
\begin{equation}
    D^j(R) = \exp(-i2\pi j/N)
\end{equation}
where $j=0,1,\cdots, n$. Since $R$ is its generator, the rest of the
representation is determined. Hence we have all the irreducible
representations of $C_n$. The table of characters may be found on page
65 of \cite{book}.

Note that since all representations are $1$-dimensional, their
products will keep being irreducible. So having two set of irreducible
representations of $C_n$ and $C_m$, will give us all the irreducible
representations of $C_{nm}$.
\subsection{Dihedral Group}
\label{sec:Dihedral-Group}
\begin{defi}[Dihedral Group $D_n$]
\nomenclature{Dihedral Group $D_n$}{\nomrefpage.}
    Dihedral group $D_n$ is the symmetric group of regular $n$-gons.
\end{defi}

\paragraph{Conjugacy Class}
There is some important difference between the case when $n$ is odd
and when $n$ is even. When $n$ is odd, the $2$-fold axis are all
linked with the vortices of the $n$-polygon. On the contrary, when $n$
is even, there is two different $2$-fold axis. The first $n/2$ axises
are those linked with the vortices, the second $n/2$ axises are those
cut through the opposite edges.

@@

In summary, this discussion shows that the self-conjugate classes in $D_n$
are:
\begin{fact}
    When $n$ is odd (Let $n=2n'+1$): The identity class, the class of
    all $2$-fold axis. The class of $\{ T^m, T^{-m}\}$, where
    $m=1,2,\cdots, n'$. (Why $m\leq n'$, think about the case when
    $n=5,n'=2$.). In total $n'+2=\frac{n+3}{2}$ conjugacy
    classes/irreducible representations. All conjugacy classes are
    self-conjugate.
\end{fact}
\begin{fact}
    When $n$ is even (Let $n=2n'$): The identity class, the two
    classes of all $2$-fold axis, as discussed before. The class of
    $\{ T^m, T^{-m}\}$, where $m=1,2,\cdots, n'$. In total
    $n'+3=\frac{n+6}{2}$. All conjugacy classes are self-conjugate.
\end{fact}

\paragraph{Invariant Subgroups}
By the distinction between edge and vortices type of $2$-fold axises
discussed before, one can see that the difference in invariant
subgroup when $n$ is odd or even. More Specifically, when $n=5$ is odd,
$D_5$ has only $C_5$ as a notrivial subgroup. But when
$n=6$ is even, it has $C_6$ (then $C_2$ and $C_3$), two copies of
$D_3$ (each formed by edge and vortex type of $2$-fold axis).

>>>>

\paragraph{Irreducible representation} (\textbf{pp.64-65 of \cite{book}})

Since it has $n$ elements and $n$ conjugacy classes, with
$\sum_{j=1}^n d_j^2 = n$, each irreducible representation is
$1$-dimensional. Let $R$ denotes the generator of rotation by
$2\pi/n$, then by calculating $D^j(R)^n=D^j(1)$, one can easily find
that for each irreducible representation $j$:
\begin{equation}
    D^j(R) = \exp(-i2\pi j/N)
\end{equation}
where $j=0,1,\cdots, n$. Since $R$ is its generator, the rest of the
representation is determined. Hence we have all the irreducible
representations of $C_n$. The table of characters may be found on page
65 of \cite{book}.

Note that since all representations are $1$-dimensional, their
products will keep being irreducible. So having two set of irreducible
representations of $C_n$ and $C_m$, will give us all the irreducible
representations of $C_{nm}$. For example:
% TODO pdf scan of C_6.

\subsection{Dihedral Group \texorpdfstring{$D_n$}{}}
\label{sec:Dihedral-Group}
\begin{defi}[Dihedral Group $D_n$]
\nomenclature{Dihedral Group $D_n$}{\nomrefpage.}
    Dihedral group $D_n$ is the symmetric group of regular $n$-gons
    for $n\geq 3$. For $n=1/2$, it is usually not defined. I found that
    in book \cite{book}, $D_2$ is isomorphic to Klein Group, $D_1$ is
    of course the trivial group.
\end{defi}

But we can have another more mathematical definition of Dihedral
group:
\marginpar{To be added}
\paragraph{Conjugacy Classes}
There is some important difference between the case when $n$ is odd
and when $n$ is even. When $n$ is odd, the $2$-fold axis are all
linked with the vortices of the $n$-polygon. On the contrary, when $n$
is even, there is two different $2$-fold axis. The first $n/2$ axises
are those linked with the vortices, the second $n/2$ axises are those
cut through the opposite edges.

@@

In summary, this discussion shows that the self-conjugate classes in $D_n$
are:
\begin{fact}
    When $n$ is odd (Let $n=2n'+1$): The identity class, the class of
    all $2$-fold axis. The class of $\{ T^m, T^{-m}\}$, where
    $m=1,2,\cdots, n'$. (Why $m\leq n'$? Think about the case when
    $n=5,n'=2$.). In total $n'+2=\frac{n+3}{2}$ conjugacy
    classes/irreducible representations. All conjugacy classes are
    self-conjugate.
\end{fact}
\begin{fact}
    When $n$ is even (Let $n=2n'$): The identity class, the two
    classes of all $2$-fold axis, as discussed above. The class of
    $\{ T^m, T^{-m}\}$, where $m=1,2,\cdots, n'$. In total
    $n'+3=\frac{n+6}{2}$. All conjugacy classes are self-conjugate.
\end{fact}

\paragraph{Invariant Subgroups} (\textbf{pp.30 of \cite{book}})
By the distinction between edge and vortices type of $2$-fold axises
discussed before, one can see that there is difference in invariant
subgroup when $n$ is odd or even. For example, when $n=5$ is odd,
$D_5$ has only $C_5$ as a notrivial subgroup. But when
$n=6$ is even, it has $C_6$ (then $C_2$ and $C_3$), two copies of
$D_3$ (each formed by edge and vortex type of $2$-fold axis).

More generally, there importnat subgroup of index for $D_n$. When $n$
is odd (let $n'2n'+1$), there are one subgroup of index $2$, i.e.
$C_n'$. When $n$ is even (let $n'=n/2$), there are three invariant
subgroup of index $2$: $C_n'$, $D_n$, and $D_n'$. Denote $n$ $2$-fold
axis as $S_j$, rotation around $n$-fold axis by $2\pi/n$ as $T$,
then:
\begin{align}
    D_n &= \{E, T^2,T^4,\cdots,T^{2n-2},S_0,S_2,\cdots,S_{2n-2}\} \\
    D_n' &= \{E, T^2,T^4,\cdots,T^{2n-2},S_1,S_3,\cdots,S_{2n-1}\}
\end{align}
Note that this list is not complete. For example, $D_6$ contains other
nontrivial invariant subgroups, listed in page 30 of \cite{book}.

\paragraph{Character Table} (\textbf{pp.66 of \cite{book}})

For $D_2$, it is isomorphic to the group
$\{1,\sigma,\tau,\sigma\tau\}$, where $\sigma$ is spatial inversion,
$\tau$ is time-reversal. Notice that this group is abelian, so it
obvious has $4$ irreducible representations. It is easy to get its
character table, which may be found on page 66, table 3.6 of
\cite{book}.

For $D_3$, it is non-abelian. As discussed previously, it has $3$
conjugacy classes, so it has 3 irreducible representations. Notice
that $D_3/C_3 \cong C_2$, so it inherets the two simple irreducible
representations of $C_2$. This gives two lines in the character
table. Also, $6-1^2-1^2=2^2$, so we have one $2$-dimensional
irreducible representation left.  Next, by fact
\ref{fact:character-of-identity}, we can determine one element in the
character for $2$-dimensional irreducible representation. The rest two
empty blanks can be filled using the orthogonal relations for the
character table.
%TODO Add the scanned example.

For general $D_n$, see the part for irreducible representations.

\paragraph{Irreducible representation} (\textbf{pp.71-74 of \cite{book}})

The general procedure is to use the invariant subgroup of index $2$
mentioned above to get the $1$-dimensional representations. And use
the relation $\sum_j m_j^2 = |D_n|$ and $\sum_j 1 = $ number of
conjugacy class to "guess" the dimensional of remaining irreducible
representations. Finally use subgroup $C_n$ to construct those
representations (induced representation) and reduce these into
irreducible ones. But how to reduce them into irreducible one is
actually the most difficult part, in my opinion.

For odd $n$, let $n=2n'+1$. We have:
\begin{align}
    \sum_{j=1}{n'+2} m_j^2 = 4n'+2
\end{align}

There is two $1$-dimensional irreducible representations, formed using
the invariant subgroup $C_{2n'+1}$. Since $4n'+2 = 2^2 n'+2$, one
guesses that the remaining ones are $n$ $2$-dimensional irreducible
representation. They are constructed (construction process could be
found in \cite{book}) to be (notice $D_{2n'+1}$ is generated by
$T_{2n'+1}$, the rotation $2\pi/(2n'+1)$, and $C'_2$, any one of the
$2$-fold reflection):
\begin{align}
    \bar{D}^{E_j}(C_{2n'+1}) &=\left( \begin{array}{cc}
    \cos(\frac{2j\pi}{2n'+1}) & -\sin(\frac{2j\pi}{2n'+1}) \\
    \sin(\frac{2j\pi}{2n'+1}) & \cos(\frac{2j\pi}{2n'+1}) \\
        \end{array} \right) \\
    \bar{D}^{E_j}(C'_{2}) &= \left( \begin{array}{cc}
                 1 & 0 \\
                 0 & -1 \\ 
                 \end{array} \right)
\end{align}
where $j=1,2,\cdots,n$.

The case for even $n$ is simialr (let $n'=n/2$). Except that here one
as three invariant subgroup of index $2$ (see dicusson about invariant
subgroup above). So it has three $1$-dimensional irreducible
representations, and $n-1$ $2$-dimensional irreducible representations.

When $n$ is odd, let $n=2n'$. (TBD)

%TODO Finish it.
